Hire Ed: Job Market Dynamics for Tenure-Track Faculty Positions in Archaeology

### 2024-11-01 analysis/paper/article.qmd
writing background

Research into

abysmal

@@

- https://www.anthropology-news.org/articles/hello-to-tristes-tropes/

>>>>

Studies of who holds tenure-track faculty positions in archaeology at US universities, shows that success in job applications is strongly determined by where applicants get their PhD from. Data from the 2014–2015 AnthroGuide publication of the American Anthropological Association shows that just ten out of over 100 US graduate programs produced over 30% of the graduates hired into tenure-track faculty positions [@speakman2018choosing]. Similarly, a network analysis of 1,918 faculty holding tenured or tenure-track positions at PhD-granting anthropology programs in the US in 2015 (including 506 archaeologists) found that the just fifteen graduate programs produced 52.8 percent of tenured and tenure-track positions [@kawaSocialNetworkUS2019]. This network analysis further found that programs with large endowments and with faculty who hold prestigious awards and are widely cited by other scholars produce the majority of tenured and tenure-track faculty in anthropology. This pattern of inequality is not unique to archaeology [@wapmanQuantifyingHierarchyDynamics2022a], but these studies draw attention to the paradox of a discipline that has many members committed to fighting social inequalities, and yet continues to reproduce its own systematic inequalities and hierarchies through hiring practices that favour an elite minority of graduates with prestigious institutional affiliations.

One way that some hiring committees are tackling these biases towards prestigious institutions is to provided detailed instructions to applicants on how to prepare their application materials so that the hiring committee can evaluate and rank the specific accomplishments of candidate across a variety of categories, rather than ranking based on prestige of the candidates' graduate schools. This has resulted in job ads that are highly prescriptive in the types of documents that applicants should submit. For example, in addition to a cover letter and CV, job ads in many fields now require applicants to submit short statements about their teaching and research record and plans. In a comparison of job ads from 1999–2000 and 2019–2020 published in _Anthropology News_ @HelloTristesTropes2021 noticed a trend toward increasingly asking for more materials from applicants. They found that twice as many 2019–2020 job ads requested writing samples compared to 1999–2000, and nearly four times as many requested statements of teaching philosophy. In their review of the 'worst job ads of 2021' @dennisWorstAnthroJob2022 reported two job ads with extreme requirements, requesting nine and thirteen documents from applicants. A notable requirement that has emerged in recent years is a diversity statement, where the applicant describes their knowledge of, prior contributions to, and future goals for advancing diversity, equity, inclusion. This requirement has been much-debated due to an experiment during 2016-2022 at several University of California campuses that used diversity statements as the first cut for selecting candidates for tenure-track faculty positions [@soucek2021diversity]. Only candidates that scored highly on their diversity statements would have the rest of their application evaluated. This experiment generated intense and widespread public debate about merits and risks of requiring and using diversity statements, drawing attention to the challenges of hiring faculty from underrepresented minorities, and resulting in some prestigious universities no longer requiring diversity statements for job applicants [@guidenAreDiversityStatements].

@@
